Adding selection to \stdpopsim: oh how the tables have turned

### 2024-11-27 analysis2_manuscript.tex
second round of edits from DRS (#33)

\section*{Abstract}
    \label{abstract}
    Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
    However, simulating realistic models that incorporate both selection and complex demographic histories
    has remained challenging, limiting our ability to benchmark statistical methods and explore theoretical predictions.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of natural selection, including background selection, selective sweeps, and arbitrary distributions of fitness effects (DFE).
    This extension maintains \stdpopsim's core principles of reproducibility and standardization while adding support
    for species-specific genomic annotations and published DFE estimates. 
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across different species and scenarios. 
    Our results highlight how selection can bias demographic inference, 
    reveal the sensitivity of DFE inference methods to model assumptions, 
    and show how genomic features like recombination rate and functional density influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible framework.

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \cite[e.g.,][]{gillespie1991causes}.
    For instance changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \cite{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus it is imperative to joingly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep{sheehan2016deep,johri2020toward}.
    However this is a challenging task from a modeling perspective \citep{johri2022prospect}.

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few}, to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting}, to
    to explore the power and limitations of statistical methods \cite[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,][]{pavlidis2010searching,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic}.
    While this is so, joint simulation of complex demography and selection
    is challenging, and requires a deep understanding of the underlying
    evolutionary processes, as well as a number of parameter choices including
    the strength of selection, the distribution of fitness effects, and the
    recombination rate, as well as a parameterized model of demography.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, sharing and comparing simulation results among studies can 
    itself be challenging, as different simulation tools may use different structures
    and parameter scalins for their models, and may not be directly comparable.
    This is a major limitation for the field, as it can make it difficult
    to assess the robustness of results, and to compare the results of
    different studies.

    % stdpopsim
    A lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirical
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Further the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Ideally, one should be able
    to model all of these features after estimates/annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background}.

@@

    % implementing selection
    \stdpopsim previously provided a wide range of models of neutral
    demographic history from roughly two dozen species \citep{lauterbur2023expanding}.
    This was accomplished through the use of a standardized interface
    and data structure that allows for curated addition of new
    species and models, and can use either
    msprime \citep{Baumdicker2022} or SLiM \citep{haller2019slim}
    as the backend engine for efficient population genetic simulation.
    To extend \stdpopsim's functionality to include models of selection, 
    we introduce an important new class representing a specfic distribution of fitness effects
    to the \stdpopsim API: the \texttt{DFE} class. 
    This class provides an interface for specifying
    a distribution of the fitness effects of new mutations, and
    is added to a \texttt{Contig} object to provide a description
    of the fitness effects of mutations along a chromosome.
    A \texttt{DFE} object can apply to a specific interval set of a chromosome,
    and can be combined with other \texttt{DFE} objects to provide, 
    a rich model of how selection may differ along a chromosome. 
    Once specified a model of selection is then run in the same way
    but currently is limited to using SLiM simulation engine, which
    is perhaps the most flexible simulation engine for modeling selection currently available.

    As with other \stdpopsim models, when incorporating selection we aim to hew as closely as possible
    to realistic genomes, and provide public annotations of specices-specific functional
    elements that can be used along with a \texttt{DFE} object to
    readily provide realistic model of selection for a specific genome and set of
    populations. A schematic of the \stdpopsim catalog is shown in Figure \ref{fig:schematic}.
    From a users perspective, one can specify a species, a portion of the genome to simulate,
    a genetic map if available, a model of demography, and a model of selection that is
    composed of a set of annotations and a \texttt{DFE} object.

@@

    The vast majority of published DFE estimates are based on a limited number of species,
    so \stdpopsim currently has implemented published DFEs from
    four species: \textit{Arabidopsis thaleana} \citep{huber2018gene}, \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining},
    humans, and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While that is so, these DFEs can be mixed and matched with the other species
    in the catalog. For instance, one could simulate a butterfly species with a Gamma DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Furthermore the user can specify a custom DFE, and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have 
    a published DFE included in the catalog. This flexibility allows for a wide range of
    models of selection to be simulated, and for the results of these simulations to be
    compared to empirical data or used as a null model for hypothesis testing.

@@

    % current numbers of DFEs / species / etc
    At this release the \stdpopsim catalog comprises 24 species (i.e. genome representations)
    for which we have implemented 28 demographic models, have 37 genetic maps, and 7 DFEs. %someone should double check the numbers
    Additions to the catalog are ongoing -- we point the reader to our previously published 
    report detailing this effort \citep{lauterbur2023expanding} and we welcome contributions from the
    community---we encourage those interested in contributing to visit https://popsim-consortium.github.io/stdpopsim-docs
    or contact one of the PopSim Consortium members for guidance.

\section*{Example Applications with Selection}
    \label{applications}
    In this section we provide a series of examples of how the new models of selection
    in \stdpopsim can be used to benchmark and compare the performance of different
    methods for population genetic inference.
    We focus on three main areas: demographic inference in the 
    single and multi-population settings, inference of the distribution of fitness effects,
    and the detection of selective sweeps, particularly in the context of background selection.

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the pairwise sequentially Markovian coalescent
    (PSMC) \citep{li2011inference}, through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by decent information \citep{santiago2020recent}. 
    While that is so, it has been well described that selection, because of its effects on
    local and more global patterns of genetic variation, can drive estimates of $N_e(t)$
    further from true census population sizes simply because selection increases the 
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}.

    Using \stdpopsim we can simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography} we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without background selection on exons. For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome. 
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we have performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted computational resources as thorough benchmarking
    isn't the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography} we show six panels, each of which shows the estimated $N_e(t)$
    from each independent sample in a five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The top row shows the results of simulations without selection, while the bottom row shows the results
    of simulations with a Gamma distributed DFE for the deleterious effects of mutations occurring exons.
    In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the bottom row are to visually reinforce that inference is done
    from a model with selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ in the more recent past for the CEU and CHB populations.
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwaiplot which uses the SFS to infer $N_e(t)$ seems to slightly underestimate $N_e(t)$
    but is reasonably close to the true $N_e(t)$ in the absence of selection.
    Likewise \smcpp seems to be more robust to the presence of selection,
    but there is some added noise to the returned estimates. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$,
    although estimates from \msmc have quite a bit of added noise even in the absence of selection.

    % DRS: I don't agree with the interpretation in the previous two sentences (starting with "Likewise").
    % MSMC is already barfing a little bit even without selection, and it is not clear to me from these
    % plots that selection makes things any worse. Also, SMC++ is doing quite well without selection so maybe
    % the minor drop in accuracy is meaningful in relative terms.
    % ADK: I edited the text-- what do you think Dan?

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without background selection on exons. The top row shows estimates of $N_e(t)$ from simulations
        without selection, while the bottom row shows estimates of $N_e(t)$ from simulations with a Gamma distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        }
    \end{figure}

    To highlight the ease of comparisons between different species and model enabled by \stdpopsim,
    we have also performed a similar benchmark using the vaquita porpoise \textit{Phocoena sinus}.
    In Figure \ref{fig:1pop-vaquita-demography} we show the results of $N_e(t)$ inference
    where we have simulated genomes under a model of vaquita demography from \textcite{robinson2022critically}.
    For these simulations we have used the genome structure and exon annotations from the vaquita genome assembly
    and a DFE from \textcite{robinson2022critically} that acts on exonic positions, however no genetic map 
    is currently available, so recombination rates are assumed to be constant across the genome.
    The demographic model is a simple two-epoch model where population size has declined in the recent past. 
    In this case we see that all methods perform well both in the absence and presence of selection,
    with a close match between the true $N_e(t)$ and the estimated $N_e(t)$ in all cases.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/estimated_Ne_t_final}
    \caption{
    \label{fig:1pop-vaquita-demography}
    Performance of methods to infer $N_e(t)$ from simulations of the vaquita porpoise genome under a single population
    model of declining population size \citep{robinson2022critically} with and without background selection on exons. 
    The top panel shows estimates of $N_e(t)$ from simulations
    without selection, while the bottom panel shows estimates of $N_e(t)$ from simulations with a Gamma distributed   
    DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
    \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
    }
\end{figure}

\section*{Estimation of the Distribution of Fitness effects}
    \label{dfe}
    Another common application of population genetic inference is to estimate the distribution of fitness effects (DFE)
    from genetic data. Here we show how the new models of selection in \stdpopsim can be used to benchmark and compare
    the performance of different methods for inferring the DFE from genetic data. Here we have performed simulations using the 
    human and vaquita porpoise models described above, with complex demographic histories and models of selection acting on exons.
    To remind the reader, the model of selection in each case is a Gamma distributed DFE for deleterious mutations.

    \begin{figure}[htbp]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_OutOfAfricaArchaicAdmixture_5R19_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
        \caption{A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a human out-of-Africa model \citep{ragsdale2019models} with a Gamma distributed DFE
        acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        The human model has three extant populations (CEU, CHB, and YRI), and parameter estimates from each
        sample are shown in different colors.
        Estimates from three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadi \citep{gutenkunst2009inferring}.}
        \label{fig:homsap-dfe.ooa}
    \end{figure}

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data-- 
    \dadi \citep{gutenkunst2009inferring}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as a current limitation of \stdpopsim
    currently is that we are not able to simulate an outgroup sequence to estimate
    divergence. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadi seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods slightly underestimated the strength of selection. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadi seems to yield overestimates.

>>>>

\section*{Abstract}
    \label{abstract}
    Natural selection is a fundamental evolutionary force that shapes patterns of genetic variation across species. 
    However, simulating realistic models that incorporate both selection and complex demographic histories
    is challenging, limiting our ability to benchmark statistical methods and explore theoretical predictions.
    \stdpopsim is a tool for facilitating the simulation of a contiually expanding catalog of population genetic models.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of natural selection, including background selection, selective sweeps, and arbitrary distributions of fitness effects (DFE).
    This extension maintains \stdpopsim's core principles of reproducibility and standardization while adding support
    for species-specific genomic annotations and published DFE estimates. 
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across different species and scenarios. 
    Our results highlight how selection can bias demographic inference, 
    reveal the sensitivity of DFE inference methods to model assumptions, 
    and show how genomic features like recombination rate and the density of functional sequence influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible framework.

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \citep[e.g.,][]{gillespie1991causes}.
    For instance changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \citep{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus it is imperative to joingly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep{sheehan2016deep,johri2020toward}.
    However this is a challenging task from a modeling perspective \citep{johri2022prospect}.

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few}, to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting},
    to explore the power and limitations of statistical methods \citep[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,][]{beaumont2002approximate,pavlidis2010searching,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic}.
    While this is so, joint simulation of complex demography and selection
    is challenging, and requires a deep understanding of the underlying
    evolutionary processes, and necessitates a number of parameter choices including
    the strength of selection and the
    recombination rate, as well as a parameterized demographic model.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, different simulation and inference tools may use different structures
    and parameter scalings for their models.

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirical
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Further the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Ideally, one should be able
    to model all of these features after estimates/annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background}.

@@

    % implementing selection
    \stdpopsim previously provided a wide range of models of neutral
    demographic history from roughly two dozen species \citep{lauterbur2023expanding}.
    This was accomplished through the use of a standardized interface
    and data structure that allows for curated addition of new
    species and models, and can use either
    msprime \citep{Baumdicker2022} or SLiM \citep{haller2019slim}
    as the backend engine for efficient population genetic simulation.
    To extend \stdpopsim's functionality to include models of selection, 
    we introduce an important new class representing a specfic distribution of fitness effects
    to the \stdpopsim API: the \texttt{DFE} class. 
    This class provides an interface for specifying
    a distribution of the fitness effects of new mutations, and
    is added to a \texttt{Contig} object to provide a description
    of the fitness effects of mutations along a chromosome.
    A \texttt{DFE} object can apply to specified intervals within a chromosome,
    and can be combined with other \texttt{DFE} objects to provide 
    a rich model of how selection may differ along a chromosome. 
    Once specified, a model of selection can then be run using \stdpopsim's standard interfaces
    but currently is limited to using SLiM simulation engine, which
    is perhaps the most flexible simulation engine for modeling selection currently available.

    As with other \stdpopsim models, when incorporating selection we aim to hew as closely as possible
    to realistic genomes, and provide public annotations of specices' functional genomic
    elements that can be used along with a \texttt{DFE} object to
    readily provide realistic model of selection for a specific genome and set of
    populations. A schematic of the \stdpopsim catalog is shown in Figure \ref{fig:schematic}.
    From a user's perspective, one can specify a species, a portion of the genome to simulate,
    a genetic map if available, a model of demography, and a model of selection that is
    composed of a set of annotations and a \texttt{DFE} object, all in an automated and user-friendly fashion.

@@

    The vast majority of published DFE estimates are based on a limited number of species,
    so \stdpopsim currently has implemented published DFEs from
    four species: \textit{Arabidopsis thaleana} \citep{huber2018gene}, \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining},
    humans \citep{huber2017determining,kim2017inference}, and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While that is so, these DFEs can be applied to other species
    in the catalog. For instance, one could simulate a butterfly species with a gamma DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Furthermore the user can specify a custom DFE, and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have 
    a published DFE included in the catalog. This flexibility allows for a wide range of
    models of selection to be simulated. %, and for the results of these simulations to be
    %compared to empirical data or used as a null model for hypothesis testing.
    % DRS: commenting that stuff out because there are obviously other use cases that could be
    % listed here (e.g. ML/ABC, and if we don't know enough about a species to do that then
    % we probably have no business doing hypothesis testing either, which could be just as
    % misleading or maybe even more so), so the simplest solution is to cut this as it is
    % not needed here.

@@

    % current numbers of DFEs / species / etc
    % DRS: are we giving this a version number?
    At this release the \stdpopsim catalog comprises 24 species (i.e. genome representations)
    for which we have implemented 28 demographic models, have 37 genetic maps, and 7 DFEs. %someone should double check the numbers
    Additions to the catalog are ongoing---we point the reader to our previously published 
    report detailing this effort \citep{lauterbur2023expanding}. We welcome contributions from the
    community and encourage those interested in contributing to visit https://popsim-consortium.github.io/stdpopsim-docs
    or contact one of the PopSim Consortium members for guidance.

\section*{Example Applications with Selection}
    \label{applications}
    In this section we provide a series of examples of how the new models of selection
    in \stdpopsim can be used to benchmark different
    methods for population genetic inference.
    We focus on three main areas: demographic inference in the 
    single and multi-population settings, inference of the DFE,
    and the detection of selective sweeps, particularly in the context of background selection.

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the pairwise sequentially Markovian coalescent
    (PSMC) \citep{li2011inference}, through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by decent information \citep{santiago2020recent}. 
    While that is so, it has been well described that selection, because of its effects on
    patterns of genetic variation, can bias estimates of $N_e(t)$
    away from true census population sizes simply because selection increases the 
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}.

    Using \stdpopsim we can easily simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography} we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without purifying selection on exons. For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome.
    % DRS: Maybe we should say something about how easy it was to do these simulations, and maybe even show
    % some example code somewhere to prove our point? Could help emphasize the utility of this resource.
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we have performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted computational resources as thorough benchmarking
    isn't the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography} we show six panels, each of which shows the estimated $N_e(t)$
    from data simulated under the five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The top row shows the results of $N_e(t)$ estimates made from simulations without selection, while the bottom row shows the results
    from simulations with deleterious mutatinos occurring in exons.
    In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the bottom row are to visually reinforce that inference is done
    from a model with selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ in the more recent past for the CEU and CHB populations.
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwaiplot which uses the SFS to infer $N_e(t)$ seems to slightly underestimate $N_e(t)$
    but is reasonably close to the true $N_e(t)$ in the absence of selection.
    Likewise \smcpp seems to be more robust to the presence of selection,
    but there is some added noise to the returned estimates. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$,
    although estimates from \msmc have quite a bit of added noise even in the absence of selection.

    % DRS: I don't agree with the interpretation in the previous two sentences (starting with "Likewise").
    % MSMC is already barfing a little bit even without selection, and it is not clear to me from these
    % plots that selection makes things any worse. Also, SMC++ is doing quite well without selection so maybe
    % the minor drop in accuracy is meaningful in relative terms.
    % ADK: I edited the text-- what do you think Dan?
    % DRS: I am still finding this description pretty inaccurate. Happy to discuss over zoom if you think
    % that might be faster.

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without purifying selection on exons. The top row shows estimates of $N_e(t)$ from simulations
        without selection, while the bottom row shows estimates of $N_e(t)$ from simulations with a gamma-distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        }
    \end{figure}
    %DRS-fig: reminder to update the legend to just say ``exons'' instead of ensembl_havana_whatever...

    To highlight the ease of comparisons between different species and models enabled by \stdpopsim,
    we have also performed a similar benchmark using the vaquita porpoise \textit{Phocoena sinus}.
    In Figure \ref{fig:1pop-vaquita-demography} we show the results of $N_e(t)$ inference
    where we have simulated genomes under a model of vaquita demography from \textcite{robinson2022critically}.
    For these simulations we have used the genome structure and exon annotations from the vaquita genome assembly
    and a DFE from \textcite{robinson2022critically} that acts on exons, however no genetic map 
    is currently available, so recombination rates are assumed to be constant across the genome.
    The demographic model is a simple two-epoch model where population size has declined in the recent past. 
    In this case we see that all methods perform well both in the absence and presence of selection,
    with a close match between the true $N_e(t)$ and the estimated $N_e(t)$ in all cases.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/estimated_Ne_t_final}
    \caption{
    \label{fig:1pop-vaquita-demography}
    Performance of methods to infer $N_e(t)$ from simulations of the vaquita porpoise genome under a single population
    model of declining population size \citep{robinson2022critically} with and without background selection on exons. 
    The top panel shows estimates of $N_e(t)$ from simulations
    without selection, while the bottom panel shows estimates of $N_e(t)$ from simulations with a gamma-distributed   
    DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
    \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
    }
\end{figure}

\section*{Estimation of the Distribution of Fitness effects}
    \label{dfe}
    Another common application of population genetic inference is to estimate the distribution of fitness effects (DFE)
    from genetic data. Here we show how the new models of selection in \stdpopsim can be used to benchmark and compare
    the performance of different methods for inferring the DFE. We performed simulations using the 
    human and vaquita porpoise models described above, with complex demographic histories and models of selection acting on exons.

    \begin{figure}[htbp]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_OutOfAfricaArchaicAdmixture_5R19_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
        \caption{A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a human out-of-Africa model \citep{ragsdale2019models} with a gamma-distributed DFE
        acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        The human model has three extant populations (CEU, CHB, and YRI), and parameter estimates from each
        sample are shown in different colors.
        Estimates from three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadi \citep{gutenkunst2009inferring}.}
        \label{fig:homsap-dfe.ooa}
    \end{figure}

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data:
    \dadi \citep{gutenkunst2009inferring}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}.
    In each case we note that the DFE is estimated from segregating sites only,
    without the use of substitution data, as a current limitation of \stdpopsim
    currently is that we are not able to simulate an outgroup sequence to estimate
    divergence. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadi seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods slightly underestimated the strength of selection. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadi seems to yield overestimates.
